Concurrent Programming in Harmony

### 2021-11-07 book/paper.tex
Added queue spec to chapter

The most common use for locks is in building concurrent data structures.
By way of example, we will first demonstrate how to build a concurrent queue.
The \texttt{queue} module will have the following API:
\begin{itemize}
\item $q = \mathtt{Queue}()$: allocate a new queue;
\item $\mathtt{put}(q, v)$: add $v$ to the tail of the queue;
\item $r = \mathtt{get}(q)$: returns $r = \mathtt{None}$
if $q$ is empty or $r = v$ if $v$ was at the head of the queue.
\end{itemize}

\noindent
See \autoref{fig:queuedemo} for a simple demonstration program that uses the queue.

We will first implement the queue as a linked list.
The implementation in \autoref{fig:queue}
uses the \texttt{alloc} module for dynamic allocation
\index{dynamic allocation}%
of nodes in the list using \texttt{malloc}() and \texttt{free}().
\texttt{malloc}($v$) returns a new memory location initialized to $v$,
which should be released with \texttt{free}() when it is no longer in use.
The queue maintains a \texttt{head} pointer to the first element in the list
and a \texttt{tail} pointer to the last element in the list.
The \texttt{head} pointer is \texttt{None} if and only if the queue is empty.
(\texttt{None} is a special address value that is not the address of any
memory location.)

@@

\begin{figure}
\begin{code}
\harmonysource{queue}
\end{code}
\caption{\harmonylink{code/queue.hny}A basic concurrent queue data structure.}
\label{fig:queue}
\end{figure}

@@

\begin{figure}
\begin{code}
\harmonysource{queuespec}
\end{code}
\caption{\harmonylink{code/queuespec.hny} Sequential queue specification.}
\label{fig:queuespec}
\end{figure}

\begin{figure}
\begin{code}
\harmonysource{qtestseq}
\end{code}
\caption{\harmonylink{code/qtestseq.hny} Sequential queue test.}
\label{fig:qtestseq}
\end{figure}

@@

As with critical sections, when testing a concurrent data structure
we need a specification for its correctness.
A good place to start is thinking about a \emph{sequential} specification
for queues.
A specification is simply a program, written at a high level of
abstraction.
\autoref{fig:queuespec} shows a sequential specification of a
queue in Harmony (exploiting some methods from the \texttt{list}
module described in \autoref{ap:list}).

>>>>

\begin{figure}
\begin{code}
\harmonysource{queuespec}
\end{code}
\caption{\harmonylink{code/queuespec.hny} A sequential specification for a queue.}
\label{fig:queueseq}
\end{figure}

\begin{figure}
\begin{code}
\harmonysource{queue}
\end{code}
\caption{\harmonylink{code/queue.hny} A concurrent specification for a queue.}
\label{fig:queue}
\end{figure}

The most common use for locks is in building concurrent data structures.
By way of example, we will first demonstrate how to build a concurrent queue.
The \texttt{queue} module has the following API:
\begin{itemize}
\item $q = \mathtt{Queue}()$: allocate a new queue;
\item $\mathtt{put}(q, v)$: add $v$ to the tail of the queue;
\item $r = \mathtt{get}(q)$: returns $r = \mathtt{None}$
if $q$ is empty or $r = v$ if $v$ was at the head of the queue.
\end{itemize}

\noindent
\autoref{fig:queueseq} shows a sequential specification for such a
queue in Harmony (exploiting some methods from the \texttt{list}
module described in \autoref{ap:list}),
but it cannot be used with threads concurrently accessing this queue.
\autoref{fig:queue} shows the corresponding concurrent specification.
See \autoref{fig:queuedemo} for a simple demonstration program that uses
a concurrent queue.

We will first implement the queue as a linked list.
The implementation in \autoref{fig:queueconc}
uses the \texttt{alloc} module for dynamic allocation
\index{dynamic allocation}%
of nodes in the list using \texttt{malloc}() and \texttt{free}().
\texttt{malloc}($v$) returns a new memory location initialized to $v$,
which should be released with \texttt{free}() when it is no longer in use.
The queue maintains a \texttt{head} pointer to the first element in the list
and a \texttt{tail} pointer to the last element in the list.
The \texttt{head} pointer is \texttt{None} if and only if the queue is empty.
(\texttt{None} is a special address value that is not the address of any
memory location.)

@@

\begin{figure}
\begin{code}
\harmonysource{queueconc}
\end{code}
\caption{\harmonylink{code/queueconc.hny} An implementation of a concurrent queue data structure.}
\label{fig:queueconc}
\end{figure}

@@

@@

As with critical sections, when testing a concurrent data structure
we need a specification for its correctness.
A good place to start is thinking about a \emph{sequential} specification
for queues.
A specification is simply a program, written at a high level of
abstraction.
\autoref{fig:queueseq} shows a sequential specification of a
queue in Harmony.
